Modelling the emergence of whisker barrels

### 2020-01-16 paper/briefpaper/brief.tex
Addresses R1's minor point

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a
reaction-diffusion style model of how extrinsic signalling gradients can
constrain the emergence of distinct fields from intrinsic cortical
dynamics. Their model defines how the density of connections $c(x,t)$ and axon
branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis
$x$, for $N$ thalamocortical projections indexed by $i$. The model was derived
from the assumption that the rates at which $a_i$ and $c_i$ grow are
reciprocally coupled. Extending the original 1D model to simulate arealization
on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. The second term on the right represents
an assumption that axon branching decreases when synaptogenesis increases, and
vice versa, i.e., that axon branches are required for synaptic connections to
form. Here $\chi_i=0$ is a placeholder.

@@

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced in our 2D implementation. Using an
elliptical boundary with $M=3$ offset guidance gradients aligned to the longer
axis, $N=5$ projections gave rise to five distinct cortical fields at
locations that preserved the topographic ordering defined by the original
$\gamma$ values. However, we found that specifying $N$ ordered areas required
$M\approx (N+1)/2$ signalling fields. As the number of distinct guidance
fields is unlikely to approach the number of barrels, further extensions were
required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
projections for limited cortical connections. To introduce axon branching
competition we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j},
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density at $t=0$;
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma},
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
center. As such, two orthogonal linear thalamic gradients were defined, from
which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41
barreloid centers recreated from \cite{haidarliu_size_2001} (Fig.\,1B).
%
A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001}, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a({\bf x},0)$, a clear Voronoi-like tesselation emerged (Fig.\,1C; Movie
S1). The Honda $\delta$ metric (see \cite{senft_mouse_1991}) confirmed that a
`good' Voronoi pattern emerged within $\approx 3000$ iterations (Fig.\,1D). A
measure of the difference in shape between real and simulated barrels revealed
a strong correspondence (see Fig.\,1D).

>>>>

\usepackage{amsmath,esint}
\newcommand*\dif{\mathop{}\!\mathrm{d}}

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a
reaction-diffusion style model of how extrinsic signalling gradients can
constrain the emergence of distinct fields from intrinsic cortical
dynamics. Their model defines how the density of connections $c(x,t)$ and axon
branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis
$x$, for $N$ thalamocortical projections indexed by $i$. The model was derived
from the assumption that the rates at which $a_i$ and $c_i$ grow are
reciprocally coupled. Extending the original 1D model to simulate arealization
on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and model
synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{j=1}^{N} c_{j}\right)[a_i]^k
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one,
connections decay at rate $\alpha$. Otherwise the connection density increases
non-linearly ($k>1$) with the density of axon branching. Axon branching is
modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of the quantity in parentheses, which is referred to as the
`flux' of axonal branching. The flux represents diffusion across the cortical
sheet, at rate $D$, and the influence of $M$ molecular signalling fields,
$\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given
thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which
may be positive or negative in order that axons may branch in the direction of
either higher or lower concentrations. The second term on the right represents
an assumption that axon branching decreases when synaptogenesis increases, and
vice versa, i.e., that axon branches are required for synaptic connections to
form. Here $\chi_i=0$ is a placeholder.

@@

First we verified that all results established by \cite{karbowski_model_2004}
for a 1D axis could be reproduced in our 2D implementation. Using an
elliptical domain, $S$, with $M=3$ offset guidance gradients aligned to the longer
axis, $N=5$ projections gave rise to five distinct cortical fields at
locations that preserved the topographic ordering defined by the original
$\gamma$ values. However, we found that specifying $N$ ordered areas required
$M\approx (N+1)/2$ signalling fields. As the number of distinct guidance
fields is unlikely to approach the number of barrels, further extensions were
required.

The term in parentheses in Eq.~\ref{eq:dc} represents competition between
projections for limited cortical connections. To introduce axon branching
competition we redefined
%
\begin{equation} \label{eq:comp}
\chi_i(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{j \ne i}^{N} a_{j}
\end{equation}
%
which reduces branching for each projection where the branches of other
projections are dense. Divisive normalization keeps the axon branch density
bounded on each iteration to the initial density at $t=0$;
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \oiint_{S}  a_i(\mathbf{x}, 0) \dif S \; \frac {a'_i(\mathbf{x}, t)} {\oiint_{S}
  a'_i(\mathbf{x}, t) \dif S}
\end{equation}
%
where prime symbols indicate the use of intermediate values computed from Eqs.~\ref{eq:dc}--\ref{eq:comp} at time $t-{\delta}t$.

The only differences between thalamic projections are their strengths of
interaction with the guidance fields, $\gamma$, hence any reliable differences
between emergent cortical fields must be due to differences in these values
only. We speculate that a projection interacts with a given ephrin field in
the cortical subplate with a strength determined by the concentration of a
similar molecule at its thalamic origin, i.e., the putative barreloid
center. As such, two orthogonal linear thalamic gradients were defined, from
which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41
barreloid centers recreated from \cite{haidarliu_size_2001} (Fig.\,1B).
%
A cortical boundary enclosing 41 corresponding barrels was traced from a
cytochrome oxidase stain from \cite{zheng_signal_2001} to define $S$, and
Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the
resulting domain, using $M=2$ linear signalling gradients aligned with the
anterior-posterior and medial-lateral axes. From random initial conditions for
$a({\bf x},0)$, a clear Voronoi-like tesselation emerged (Fig.\,1C; Movie
S1). The Honda $\delta$ metric (see \cite{senft_mouse_1991}) confirmed that a
`good' Voronoi pattern emerged within $\approx 3000$ iterations (Fig.\,1D). A
measure of the difference in shape between real and simulated barrels revealed
a strong correspondence (see Fig.\,1D).
